### 2015-02-06 thesis-proposal/problem.tex
Updated proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}
-- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
%Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries. % In such a system, new dialects and revisions of the system would rarely be necessary. %Indeed, a preponderence of new dialects is clear evidence against claims of generality.  %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice. %Library implementations are not yet satisfying.

We observe that new system dialects are often motivated by a desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over. %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
Introducing mechanisms that decentralize control over these features could decrease the need for new dialects and thus increase generality. But this must done with care, because giving away too much control could weaken important metatheoretic guarantees. Moreover, such  mechanisms must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them. Having to manually resolve conflicts between libraries or reestablish important properties of a program whenever a new library is imported introduces undesirable complexity. %It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}). For programmers, 
%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).

In this thesis, we aim to design mechanisms that handle these issues by embracing the already widely accepted notion that \emph{type constructors} (or more concisely, \emph{tycons}) can serve as the fundamental organizational unit of a programming system. We show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to user-defined functions associated with type constructors, together forming what we call \emph{active type constructors}. This permits library-based realizations of features that today would need to be built into a system by its central designers. By layering these mechanisms over a fixed typed internal language and constraining these functions appropriately, we retain key  metatheoretic properties and arrive at powerful modular reasoning principles. % while maintaining key metatheoretic properties and modular reasoning principles. % By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem. %By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system. 
%We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
%By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly. %We call user-defined types that introduce new features into the system in this way \emph{active types}.  % also highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the use cases we seek to address more concrete, let us begin with a simple example that we will return to throughout this proposal. \emph{Regular expressions} are commonly used to capture patterns in strings (e.g. motifs in DNA sequences) \cite{Thompson:1968:PTR:363347.363387}. Programmers who work with regular expressions would benefit from features like these:

\begin{enumerate}
\item \textbf{Concrete syntax for pattern literals}. An ideal syntax would permit programmers to express patterns in the concise, conventional manner. For example, consider a bioinformatics  application: the BisI restriction enzyme cuts DNA when it sees the motif \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} is any base. We would want to express such patterns with syntax like this, using curly braces to splice one pattern into another and parentheses to delimit captured groups:
\begin{lstlisting}[numbers=none]
let N : Pattern = <SURLA|T|G|CEURL>
let BisI : Pattern = <SURLGC({EURLNSURL})GCEURL>\end{lstlisting}
%The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
Malformed patterns would result in intelligible {parser} errors.
\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained. It might statically ensure that: 
	\begin{enumerate}
	\item only patterns, or properly escaped strings, are spliced into a pattern literal, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
	\item out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}
	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis,HosoyaVouillonPierce2000ICFP}
	\end{enumerate}
%When an error is found, an intelligible error message is provided.
%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} to support syntax highlighting, consulting relevant documentation, interactive testing and pattern extraction from example strings.
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today provides built-in support for all of the features enumerated above in their strongest form, so library providers must leverage  general-purpose mechanisms. %Unfortunately, it is impossible to completely define the syntax and the semantics referenced above using contemporary mechanisms, so library providers need to compromise.

The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time. This provides only a partial approximation to feature 1 due to clashes between string escape sequences and regular expressions (parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}). Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}. It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.  Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g. syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.

A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g. as exposed by functional datatypes or objects), or using an abstract type. This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax). Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to this  in Sec. \ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.

None of these approaches address the important issue of tool support (feature 3). A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g. \cite{_txt2re:_????}). We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}. Unfortunately, evidence suggests that tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}. We have found that they are also less usable than editor-integrated tools %(e.g. \cite{IntelliJRegexp}) 
because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}. Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding. For these reasons,   we see editor services as within the scope of an abstraction specification.
\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools. (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
\label{approaches}
%\vspace{-10px}
\end{figure}

In situations, like these, where a library-based approach is not satisfying, providers must take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects. That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool. For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}). A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach. A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same. This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a. Although often justified as ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation and such a construction gives us no rigorously justifiable  reason to believe that it can safely and naturally coexist with others specified or implemented in the same way. That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}. There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there can be serious interference and safety issues, as we will discuss at length throughout this thesis.
% More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

%\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features. This makes adoption more costly when these incidental features are  not desirable or are insufficiently developed (``toy languages''), or when the features bundled with a different language or tool are simultaneously desirable.

Recent evidence indicates that this is one of the major barriers preventing research on programming languages from being driven into practice. For example, developers prefer high-level language-integrated parallel programming abstractions that provide  strong semantic guarantees when all else is equal \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level. This is problematic because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}. Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom or research setting. Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means. Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.% This intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}:  ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

%\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden. 
%\item

%Clients seem to prioritize the ability to choose different features for different portions of an application. 
%If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
%Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.

\paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g. a team goes through the trouble of implementing \verb|G+R+P| by reading papers about \verb|G+R| and \verb|G+P| and disentangling orthogonality-related issues), there remains a problem at any interface between  components written using a different combination of features. An interface that  externally exposes a specialized construct particular to one language (e.g. a function that requires a quantity having a particular unit of measure) cannot necessarily be safely and naturally consumed from another language (e.g. a parallel programming language). Tool support is also lost when calling into different languages. %We call this the \emph{interoperability problem}. % programs written by clients of a certain collection of features cannot always interface with programs written by clients of other features  in a safe, performant and natural manner.

One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages. Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy. As indicated in Figure \ref{approaches}a, this only enables interoperability in one direction. Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not, unless these languages are only trivially different from the common language.

%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. 
As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET. This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries. Indeed, guidelines for F\# suggest that exposing functions that operate over values having units of measure, datatypes or tuples is not recommended when a component ``might be used'' from another language \cite{syme2012expert} because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g. pattern matching) and syntax that F\# includes. SML.NET prohibits exposing such types at component boundaries altogether. Moreover, it also cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML). 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
%In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
%\end{itemize}

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Introduction}\label{motivation}
% \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
% and generality in language design is a serious unsolved 
% problem.} -- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
Well-designed general-purpose programming languages are remarkably stable: they rarely need to be revised or forked into dialects, because programmers can express the constructs that they need just as well in orthogonal libraries. %Instead, . . %Indeed, a preponderence of new dialects is clear evidence against claims of generality.  %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
Abstraction mechanisms informed by type theory, like  the ML module system \cite{MacQueen:1984:MSM:800055.802036}, have put this ideal within reach. 
Yet despite their expressive power, new {dialects} of languages like ML continue to arise. Why is this?

% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed. %This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.   %Library implementations are not yet satisfying.
%Why would a well-informed programmer consider forming a new dialect? 
Perhaps the most common justification is that the \emph{syntactic cost} of a desirable construct is not  preserved when it is expressed using general-purpose mechanisms. %Syntactic cost is often assessed qualitatively \cite{green1996usability} (though quantitative metrics can also be defined). 
For example, the semantics of regular expressions can be expressed using abstract data types (which in ML are a mode of use of the module system), but as we will detail in Sec. \ref{sec:syntax}, this presents some syntactic difficulties. For programmers in problem domains  where regular expressions are common, e.g. bioinformatics, it is tempting to design a ``domain-specific'' dialect that differs only in that it includes derived syntax (i.e. ``syntactic sugar'') for regular expression patterns. There is no shortage of tools that aid in the construction of such syntactic dialects (e.g. Camlp4 \cite{ocaml-manual}, SugarJ \cite{erdweg2011sugarj}, SugarHaskell \cite{erdweg2012layout} and many others).%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).

Somewhat more rarely, dialects are motivated by the need to extend a general-purpose language with new derived semantic constructs, i.e. constructs that cannot be expressed as purely syntactic sugar, but where the general-purpose language is a suitable translation target. As a minimal example, consider System \textbf{F}, i.e. the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{pfpl}. If we wanted to express record types in \textbf{F}, we could not do so simply with derived syntax. We could, however, develop a dialect that included the type and term operators relevant to records, specifying their static semantics directly but their dynamic semantics by translation to the polymorphic lambda calculus (using a standard Church-style encoding).\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.

% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators. However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone. \todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations. \todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.

The problem with designing new language dialects in these situations is that a program written in one dialect cannot safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the two languages target a common intermediate language,  
To do so would require combining the two dialects into a single language, but there is no systematic method to do so, much less one that guarantees that important metatheoretic properties established about each dialect in isolation will be conserved. As a simple example, two dialects might be known to have an unambiguous concrete syntax in isolation, but if the syntax is na\"ively combined, ambiguities could easily arise. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
Without strong modular reasoning principles to rely upon, it is difficult to construct large software systems and ecosystems. 
%This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
The only reasonable approach for software engineers who must program ``in the large'' is to eschew constructs housed in dialects and settle on a single common language, where they can  rely on the guarantees provided by its type and module system. %The designers of these languages may slowly adopt the most generally useful new ideas. 
This has left many useful constructs for programming ``in the small'' obscure.% It is safe to say that in contemporary times, language dialects should be considered harmful. % It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.

\section{Contributions}
Our aim in this thesis is to make dialect formation less frequently necessary by designing language mechanisms that give library providers the ability to express derived syntactic and semantic constructs from within the language, in a manner that ensures that these constructs can be reasoned about separately and composed arbitrarily.

%Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself. %Defining a new construct then corresponds to the usual notion of language extension. We will make this intuition more precise later.

%in languages without such mechanisms, expressing such constructs would require extensions to the language itself (a better qualifier might thus be \emph{extension-mitigating}). %We will make these notions more precise as we continue. %Thus, these mechanisms precisely capture the notions of language extension and language composition.

%The mechanisms we propose are designed to decrease the number of constructs that need to be built in \emph{a priori} by a language designer. However, there are still some basic syntactic and semantic decisions that must unavoidably be made.

To provide a coherent vehicle for these mechanisms, we will introduce a new general-purpose language called Wyvern. Let us briefly summarize its organization. Syntactically, Wyvern has a layout-sensitive textual concrete syntax (i.e. indentation is meaningful). This choice is not fundamental to our proposed mechanisms, but it will be useful for a class of examples that we will discuss later.  Semantically, Wyvern consists of a \emph{module language} and a \emph{core language}. The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}. The core language (i.e. the language of types and expressions) is split into an \emph{external language} (EL) and an \emph{internal language} (IL). The semantics of EL constructs are specified by a type-directed translation to the much simpler IL. Notionally, this can be thought of as simply a shifting of the first stage of a type-directed compiler into the language specification \cite{tarditi+:til-OLD}. The internal language we will use is the polymorphic lambda calculus with binary product and sum types and recursive types (features like state and exceptions could also be included, but for simplicity, we will omit them) \cite{pfpl}. %Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}. % (for the purposes of this work).

We will begin by describing Wyvern's mechanisms for expressing new forms of derived concrete syntax, assuming for the moment that the EL is otherwise specified in a standard way, in Section \ref{sec:syntax}. We then turn in Section \ref{sec:semantics} to a mechanism for introducing new semantic constructs -- in particular, new base types and operators -- into the EL. Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Wyvern be realized as modes of use of this mechanism. In both cases, the mechanisms can be broadly understood as novel forms of \emph{static metaprogramming}. Users define each new syntactic and semantic construct by writing code that generates terms and types. This code is  evaluated statically, in the course of typechecking EL terms.

%Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.

For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Wyvern be expressed in libraries. To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties. The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined syntactic and semantic constructs can be reasoned about separately and then used together in any combination. These justify our classification of Wyvern as a \emph{modularly extensible programming language}.

%Designing a modularly extensible language is a challenge because giving away too much control can easily make reasoning about program behavior difficult and lead to problems that arise only when extensions are used at the same time. For example, a library provider given free reign over syntax and semantics could introduce parsing conflicts, violate {type safety}, and  interfere with other extensions. We aim to design mechanisms that eliminate these possibilities, while providing a degree of expressive power sufficient to express a variety of constructs that are, or would need to be, built directly in to today's languages. %ll constructs that can be defined using a mechanism we will introduce can be reasoned about separately and composed arbitrarily.% With these mechanisms, we anticipate dialect formation being far less common. %Indeed, all of these problems have plagued previous  designs.

% Common dialects of ML that suffice for \cite{harper1997programming}. %We will .

   %Our challenge will be to design mechanisms that achieve these goals while also being powerful enough to express features that today require dialects.%Our challenge will be to design such extension mechanisms, while still retaining significant expressive power -- we wish to permit library-based expression of a range of features that today are, or would need to be, built into a system by its designers.%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}). For programmers, 
%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).

%So our goal will be to design extension mechanisms that handle these metatheoretic and composition-related issues while still  permitting library-based expression of a range of features that today are, or would need to be, built into a system by its designers. %More specifically, we will show how to delegate control, in a methodical manner, to \emph{static functions}, i.e. user-defined functions written in a \emph{static language}. By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles. %By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features. % while maintaining key metatheoretic properties and modular reasoning principles. % By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem. %By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system. 
%We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
%By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly. %We call user-defined types that introduce new features into the system in this way \emph{active types}.  % also highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.

%The result of this work will be the design of a \emph{modularly extensible programming language} called Wyvern.   and end with a brief foray into editor services in Sec. \ref{sec:editor-services}.\todo{fix sections}

\section{Modular Derived Syntax}\label{sec:syntax}
Most contemporary computer programming languages specify a textual concrete syntax. Because it serves as the language's human-facing user interface, it is common practice to include  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally. % (i.e. considering cognitive dimensions \cite{green1996usability}). 
For example,  list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}. Many languages go further, building in syntax associated with other types of values, like vectors (SML), arrays (Ocaml), commands (Haskell) and syntax trees (Scala).  %The desugaring from the latter to the former is specified by the language itself. %Typically, the language designer controls what forms of derived syntax are built in to the language.

Rather than privileging  particular data structures, Wyvern instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own. %Lists need no special consideration from the language specification.
%The purpose of this section is to motivate and then introduce Wyvern's syntax extension mechanisms. %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
To motivate our desire for this level of syntactic control, we begin in Sec. \ref{sec:examples} with the example of regular expression patterns expressed using abstract types, showing how the usual workaround of using strings to introduce patterns is not ideal. We then survey existing syntax extension mechanisms in Sec. \ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs. In Sec. \ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they resolve these issues. We conclude in Sec. \ref{sec:syntax-timeline} with a timeline for remaining work.

\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
Let us begin by taking the perspective of a \emph{regular expression} library provider. Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}. We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Wyvern supports these as case types, which are analagous to datatypes in ML:

\begin{lstlisting}[numbers=none]
casetype Pattern
  Empty
  Str of string
  Seq of Pattern * Pattern
  Or of Pattern * Pattern
  Star of Pattern
  Group of Pattern
\end{lstlisting}

However, there are some reasons not to expose this representation of patterns directly to clients. First, patterns are usually identified up to congruence. For example, $\textbf{seq}(\textbf{empty}, p)$ is congruent to $p$. It would be useful if congruent patterns were indistinguishable from the perspective of client code. Second, it can be useful for performance reasons to maintain additional data alongside patterns (e.g. a corresponding finite automata here) without exposing this ``implementation detail'' to clients. Indeed, there are many ways to implement regular expression patterns, each with different performance trade-offs. For these reasons, a better approach is to define the following \emph{module type} (a.k.a. \emph{signature} in SML), where the type of patterns, \lstinline{t}, is held abstract. The client of any module \lstinline{P : PATTERN} can then identify patterns as terms of type \verb|P.t|. 
Notice that it exposes an interface otherwise isomorphic to the one available using a case type:

\begin{lstlisting}[deletekeywords={case},numbers=none]
module type PATTERN
  type t
  val Empty : t
  val Str : string -> t
  val Seq : t * t -> t
  val Or : t * t -> t
  val Star : t -> t
  val Group : t -> t
  val case : (
    'a -> 
    (string -> 'a) ->
    (t * t -> 'a) ->
    (t * t -> 'a) ->
    (t -> 'a) ->
    (t -> 'a) -> 
    'a)
\end{lstlisting}

\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be used directly in all but the most trivial examples, so patterns are conventionally written using a more concise concrete syntax. For example, the concrete syntax \lstinline{A|T|G|C} corresponds to abstract syntax with the following encoding:
\begin{lstlisting}[numbers=none,mathescape=|]
P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
\end{lstlisting}

% \begin{figure}
% \begin{center}
% \includegraphics[scale=.55]{regex-palette.pdf}\\
% $\Downarrow$ \text{(pressing \textbf{Enter})}\\
% \includegraphics[scale=.55]{regex-code-generated.pdf}
% \end{center}
% \vspace{-20px}
% \caption{An example of type-specific editor services, here shown for Java.}
% \label{fig:regex-palette}
% %\vspace{-10px}
% \end{figure}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools.

To encode the concrete syntax of patterns, regular expression library providers usually provide a utility function that converts strings to patterns. Because there may be many pattern implementations, the usual approach is to define a parameterized module (a.k.a. \emph{functor} in SML) defining such utility functions generically, like this:

\begin{lstlisting}[numbers=none]
module PatternUtil(P : PATTERN)
  fun parse(s : string) : P.t
    (* ... pattern parser here ... *)
\end{lstlisting}
This allows the client of any module \lstinline{P : PATTERN} to use the following definitions:
\begin{lstlisting}[numbers=none]
module PU = PatternUtil(P)
let pattern = PU.parse
\end{lstlisting}
to construct patterns like this:
\begin{lstlisting}[numbers=none]
pattern "SSTRA|T|G|CESTR"
\end{lstlisting}
 %Again, none of our goals are comprehensively achieved.
This approach is imperfect for several reasons:
\begin{enumerate} 
\item Our first problem is syntactic (or, one might say, aesthetic): string escape sequences conflict syntactically with pattern escape sequences. For example, the following will not be well-formed:
\begin{lstlisting}[numbers=none,mathescape=|]
let ssn = pattern "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
\end{lstlisting}
In fact, when compiling an analagous term using SML of New Jersey (SML/NJ), we encounter the rather puzzling error message \verb|Error: unclosed string|. Using \verb|javac|, we encounter \verb|error: illegal escape character|. In a small lab study, we observed that this error was common, and nearly always initially misinterpreted by even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}. The workaround is to  double backslashes:
\begin{lstlisting}[numbers=none]
let ssn = pattern "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
\end{lstlisting}
